Real-Time Estimation of COVID-19 Infections via Deconvolution and Sensor Fusion

### 2022-01-21 paper/paper.tex
Revision to paper
- Address all comments from Referee 1
- Copyedits throughout
- Give Gostic et al more credit for the relevance of their paper
- Use a linear relationship between R and r, for simplicity, and
  not the log-change-ratio model from Bettencourt and Ribeiro

To give some background, in the early days of the pandemic, a handful of
non-gonvermental groups such as JHU CSSE \citep{Dong:2020} (and also the COVID
Tracking Project, the New York Times, and USAFacts) became known as the most 
trustworthy sources for aggregate public health reporting data on COVID-19 in
the U.S. They were founded around the idea of scraping COVID-19 data published
daily on dashboards that are run by local public health authorities (such as
state and county departments of public health), which, at the time, provided
more accurate and timely data than federal health authorities (probably due to
unrecoverable failures at one or more points along the reporting pipeline). In
fact, not only in the early days of the pandemic, but throughout, the data
published by these groups has been invaluable for decision-makers, modelers,
data journalists, and the general public; for example, in the U.S.\ COVID-19
Forecast Hub \citep{ForecastHub}, an open, community-driven repository of
forecasts that serves as the official source for forecasting communications by
the U.S.\ CDC, data from JHU CSSE remains the gold standard for COVID-19 case
and death forecast evaluation.

@@

In what follows in this paper, we use the CDC line list to estimate a delay
distribution between symptom onset and report dates, and then use this delay  
distribution to deconvolve daily numbers of new case reports published by JHU to
estimate daily numbers of new symptomatic infections. Furthermore, we train
models that track historical trends between past infection estimates and
auxiliary signals---based on COVID-related outpatient visits, self-reported
symptoms from online surveys, and Google searches related to COVID
symptoms---and fuse together predictions from these models to improve
the accuracy and robustness of our estimates of new infections for the most
recent 10 days (where deconvolution is particularly challenging). An
illustration is given in Figure~\ref{fig:deconv_demo}.

@@

@@

The latent setting exhibits another degree of diversity within itself. In our
work, we target symptomatic COVID-19 infections, which, to be perfectly clear,
is a latent data stream. Another example along similar lines is
\citet{Goldstein:2009}, who estimate influenza infection incidence via Bayesian
deconvolution of mortality data. Meanwhile, other authors might view inferring 
latent infections as just a stepping stone toward ultimately estimating the 
effective reproductive number $R_t$, a key epidemic parameter. Central
contributions to the methodology on $R_t$ nowcasting include:
\citet{Wallinga:2004}, who develop a simple model that uses case counts and an 
estimate of the serial interval; \citet{Bettencourt:2008}, who derive a local
approximation to the SIR model; and \citet{Cori:2013, Thompson:2019}, who build 
on the renewal equation using Bayesian modeling. For a nice review of these
approaches, see \citet{Gostic:2020}.

The aforementioned methods have been applied and extended to build systems for
real-time $R_t$ nowcasting during the COVID-19 pandemic by \citet{Abbott:2020,
  rtlive, Chitwood:2021}. A key difference between these approaches and ours is 
that they infer infections through forward-filling: loosely speaking, they
convolve forward a candidate estimate of infections, obtain feedback by
comparing the result to measured cases, and iterate to refine estimates. This
can be effective given accurate prior knowledge, but of course it can be hard to
judge the accuracy of prior knowledge in practice. We take a more flexible,
non-model-based approach and estimate infections by direct deconvolution. We
also focus on fusing in auxiliary sources of information to improve real-time
accuracy and robustness. We remark that nowcasts of $R_t$ could be inferred as a
by-product of our infection nowcasts (based on a local approximation to the SIR
model), which we describe in the discussion.

Finally, deconvolution has been extensively studied for many years in many
fields, notably signal and image processing, where deconvolution is sometimes
called deblurring. As an inverse problem, deconvolution is ill-posed in
settings in which the convolution operator is not known exactly or observations are
made with noise \citep{Oppenheim:2017}. Longstanding approaches to overcome
mis-specification here traditionally involve regularization, 
classically, Wiener deconvolution \citep{Wiener:1964}, which stabilizes the
inversion using an estimated signal-to-noise ratio. Alternative approaches use
familiar regularization techniques like $\ell_1$ and $\ell_2$ penalities
\citep{Taylor:1979, Debeye:1990}. Most related to our work is deconvolution
using total variation regularization, first proposed by \citet{Rudin:1994},
which has since become a central tool in signal and image processing.

@@

Roughly speaking, we study these particular indicators (ordered roughly from
``late'' to  ``early'') because conceptually they reflect data measurements
that would be made at some period of time in between infection onset and case
report to a public health authority, and therefore would be relevant in
inferring latent infection rates. More information on these indicators and
their underlying data sources is given in \citet{Reinhart:2021}. For more
information on CTIS in particular, we refer to \citet{Salomon:2021}. For a
study of how these and similar indicators can contribute to COVID-19 forecasting
models, we refer to \citet{McDonald:2021}.

@@

In this paper, we consider (as described in Section~\ref{sec:sensor_fusion})
various methods for combining these estimates into single estimate of the 
infection rate, which we will call \emph{sensor fusion} methods. Broadly
speaking, sensor fusion is a form of ensembling, which is used extensively 
in predictive modeling in statistics and machine learning, as it can often help
improve both accuracy and robustness. In our particular application, the sensors
themselves are constructed from data streams operating outside of traditional 
public health reporting, which itself contributes an important angle in terms of
robustness in the estimation of latent infection rates (as these are tied to  
public health reports via a convolutional model).

@@

When we say above that nowcasts are made using data that would have been
available \emph{as of} a given nowcast date $t$, we mean that we adhere not to
only the real-time availability (latency) of signals at $t$, but also the 
\emph{version} of the data published at $t$---simply put, imagine that we
``rewind'' the clock to time $t$ and query the API to receive the data that
would have been returned then. This is possible becausse the COVIDcast API
records and provides access to all historical versions of data, as described in 
\citet{Reinhart:2021}. As epidemic data is often subject to revision, if we
train and evaluate models on ``finalized'' data (that would have been available
only at a much later time point) then this can lead to overly optimistic
(inaccurate) conclusions about real-time model performance; see, e.g.,
\citet{McDonald:2021}.

Further, it is worth noting that reported case data from JHU is available at a
1-day lag, and we assume that there is at least another 1-day lag between
symptom onset and case report (explained in
Section~\ref{sec:delay_distribution}). Hence through real-time deconvolution
alone we would be able to make nowcasts at a 2-day lag at the earliest. Making
nowcasts at a 1-day lag is enabled by sensor fusion, using auxiliary signals
with 1-day latency (explained in Section~\ref{sec:leverage_aux}). In this sense,
sensor fusion is able improve not only nowcasting accuracy, but also latency,
and buys us 1 extra day.

@@

We produce nowcasts at the county resolution, but for computational purposes, we
restrict our attention to the the 200 U.S.\ counties with the highest population
(according to the 2019 Census). We additionally produce estimates for each 
of
the 50 U.S.\ states.

Some of the methodology that we use for sensor fusion---in particular, the
Kalman-filter-inspired method in \citet{Farrow:2016}---requires a
geographical hierarchy, thus using the remaining $\approx$ 2500 U.S.\ counties
we aggregate these within each state to create ``rest-of-state'' jurisdictions,  
and make estimates for these as well, for the purposes or maintaining such a
hierachy.

@@

Lastly, under an assumption that the confounding acts as a multiplicative bias
that changes slowly over time, our real-time infection rate
estimates---themselves subject to confounding by construction, as explained
above---can be post-processed to derive real-time \emph{approximately
unconfounded} estimates of $R_t$, which is the average of number of secondary
infections generated by a COVID-19 infection at time $t$. This is also described
in the discussion.

@@

Assumption~\ref{asm:geo_invar} is the most noteworthy and troublesome. We do
\emph{not} believe it to be true that different locations actually have
identical patterns of delay between infections and case reports; conversely, we 
expect there to be a considerable amount of variability between locations in
this regard. While we do allow the delay distribution to change over time (see
Figure~\ref{fig:line_list_time} for evidence for the importance of this), we 
consider Assumption~\ref{asm:geo_invar} to be a weakness of our work. However,
the \emph{data is simply not there} in the CDC line list to warrant estimation
of location-specific delay distributions (see Figure~\ref{fig:line_list_state}), 
thus we resort to estimating a nation-wide delay distribution.

@@

\begin{figure}[tb]
\centering	
\includegraphics[width=0.95\linewidth]{./figures/right_truncation_illustration.pdf}
\caption{Illustration of right truncation with a delay distribution of length 3
  (which is taken to be stationary for simplicity). At the nowcast time $t$,
  some ``part'' of the latent signal $x_t$ will appear in $y_{t+1},y_{t+2}$;
  likewise, some ``part'' of $x_{t-1}$ will appear $y_{t+1}$.}   
\label{fig:right_truncation}
\end{figure}

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=.315\linewidth]{./figures/ny_tf.pdf}
\includegraphics[width=.315\linewidth]{./figures/ny_ntf.pdf}
\includegraphics[width=.315\linewidth]{./figures/ny_tapered_ntf.pdf}
\caption{Comparison of boundary behavior for real-time deconvolution in New 
  York, displayed for a sample of nowcast dates. The black dashed line indicates 
  finalized infections, estimated roughly three months after June 1, 2021.}
\label{fig:boundary_comparison}
\end{figure*}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.825\linewidth]{./figures/tapered_smoothing_03_small_square.pdf}
\caption{Comparing regularization approaches by MAE for nowcasting. Each
  additional regularization method gives a huge improvement on trend filtering,
  and the biggest improvement is from using natural trend filtering constraints
  and a tapered smoothing penalty combined.}      
	\label{fig:extra_reg}
\end{figure}

@@

The effective reproductive number $R_t$, the average number of secondary
infections generated from a single primary infection at time $t$, is a useful
and interpretable parameter that reflects the dynamics of epidemic growth in a  
population. In the standard SIR model, the effective reproduction number at 
time $t$ can be approximated using the number of new infections at $t$  
and $t+1$, denoted by $N_t$ and $N_{t+1}$, via the relationship: 
\begin{equation}
\label{eq:log_change_ratio}
\log \frac{N_{t+1}}{N_t} \approx \gamma (R_t - 1)  
\end{equation}
where $\gamma$ denotes the recovery rate in the SIR model. This approximation 
is based on a simple Taylor series argument, and is used by
\citet{Bettencourt:2008} and other authors, though for completeness we give
a derivation for \eqref{eq:log_change_ratio} in
Appendix~\ref{app:log_change_ratio}.

Though the log-change-ratio approximation \eqref{eq:log_change_ratio} for $R_t$
is well-known in the literature on mathematical modeling of epidemics, its use
in the presence of confounding seems to be underexplored and potentially
undervalued. For simplicity, we will explain the idea in the context of a single
location. Suppose, as before, that a fraction $a_t$ of infections go untested at
time $t$. Then we have $N_t = b_t x_t$, where $x_t$ is the number of new
infections at time $t$ that show up in case reports (the focus of this paper)
and $b_t = 1/(1-a_t)$. Moreover, from \eqref{eq:log_change_ratio}, we have
$$
R_t \approx 1 + \frac{1}{\gamma} \log \frac{b_{t+1} x_{t+1}}{b_t x_t}  
\approx 1 + \frac{1}{\gamma} \log \frac{x_{t+1}}{x_t},
$$
where the last approximation is motivated by an additional assumption the 
untested fraction varies slowly over time (so that $b_{t+1}/b_t \approx
1$). The last display shows that estimates of $x_t$ (as we produce in this
paper), even though $x_t$ is itself confounded due to a lack of universal
testing, can produce \emph{approximately unconfounded} estimates of $R_t$. This
is true both in the retrospective and real-time sense, and will be the topic of
future study.

@@

\section{Log-Change-Ratio Approximation for $R_t$} 
\label{app:log_change_ratio}

We consider the standard susceptible-infected-recovered (SIR) model
\citep{Kermack:1927} with transmission rate $\beta$ and recovery rate
$\gamma$. Let $S(t),I(t),R(t)$ denote the fraction of susceptible, infected,
recovered individuals, respectively, at time $t$, in some underlying
population. Then the SIR model evolves according to the continuous-time dynamics
(subject to some initial conditions):
\begin{align*}
S'(t) &= -\beta S(t) I(t) \\
I'(t) &= \beta S(t) I(t) - \gamma I(t) \\
R'(t) &= \gamma I(t).
\end{align*}

The effective reproduction number in this model is defined as $R_t = S(t) \beta
/ \gamma$. We can interpret the $R_t$ as being determined by
the slope of a natural first-order approximation to the curve $\log I(t)$ at
time $t$. That is, looking back at the differential equation for $I(t)$, if we
forget about the fact that $S(t)$ actually depends on $t$, this ``looks like'' 
$I'(t)$ is a ``constant'' times $I(t)$ (that ``constant'' being $\beta S(t) -
\gamma$), which suggests that for $t$ near a point $t_0$,   
\begin{align}
\nonumber
I(t) &\approx I(t_0) \exp\big( (\beta S(t) - \gamma) (t-t_0) \big) \\ 
\label{eq:inf_approx}
&= I(t_0) \exp\big(\gamma (R_t - 1) (t-t_0)\big).
\end{align}

Now let $T(t)$ denote the total fraction of individuals who have ever been
infected at time $t$ (note, $T(t) = 1-S(t)$). Then this evolves according to the 
dynamics: 
$$
T'(t) = \beta S(t) I(t).
$$
Consider the change in total infecteds from time $t$ to $t+1$, also called the
\emph{incidence proportion} over this period, denoted $\Delta T(t+1) = T(t+1) - 
T(t)$. Using this as a discrete approximation to the derivative (backward
difference) of the total infecteds curve, 
\begin{align*}
\Delta T(t+1) &\approx \beta S(t+1) I(t+1) \\
&\approx \beta S(t) I(t) \exp\big(\gamma (R_t - 1)\big),
\end{align*}
where in the second line we used the local-linear approximation to log-infecteds
curve from \eqref{eq:inf_approx}, as well as a local-constant approximation to
the susceptible curve (which underlies this local-linear approximation in the
first place). By the same logic,  
$$
\Delta T(t) \approx \beta S(t) I(t),
$$
and thus 
$$
\frac{\Delta T(t+1)}{\Delta T(t)} \approx \exp\big(\gamma (R_t - 1)\big). 
$$
In other words, from the ratio $\Delta T(t+1) / \Delta T(t)$, we can estimate
$R_t$, provided that we already know (or can estimate) $\gamma$. This
establishes the claim in \eqref{eq:log_change_ratio} by taking logs of each
side, and noting that if $N_t$ is the number of new infections at each $t$, then
$N_{t+1} / N_t = \Delta T(t+1) / \Delta T(t)$ (because the total number of
individuals in the population cancels in the ratio).
\end{appendix}

>>>>

To give some background, in the early days of the pandemic, a handful of
non-gonvermental groups such as JHU CSSE \citep{Dong:2020} (and also the COVID
Tracking Project, the New York Times, and USAFacts) became known as the most 
trustworthy sources for aggregate public health reporting data on COVID-19 in
the U.S. They were founded around the idea of scraping COVID-19 data published
daily on dashboards that are run by local public health authorities (such as
state and county departments of public health), which, at the time, provided
more accurate and timely data than federal health authorities (probably due to
unrecoverable failures at one or more points along the reporting pipeline). In
fact, not only in the early days of the pandemic, but throughout, the data
published by these groups has been invaluable for decision-makers, modelers,
journalists, and the general public; for example, data from JHU CSSE remains the
gold standard for COVID-19 case and death forecast evaluation in the COVID-19
Forecast Hub \citep{ForecastHub}, a community-driven repository of forecasts
that serves as the official source for forecasting communications by the U.S.\
CDC.

@@

In what follows in this paper, we use the CDC line list to estimate a delay
distribution between symptom onset and report dates, and then use this delay
distribution to deconvolve daily numbers of new case reports published by JHU
CSSE to estimate daily numbers of new symptomatic infections. Moreover, we train
models that track historical trends between past infection estimates and
auxiliary signals of COVID-19 activity from Delphi's COVIDcast project
\citep{Reinhart:2021}, and we fuse together predictions from these models in
order to improve the accuracy and robustness of our estimates of new infections
for the most recent 10 days (where deconvolution is particularly
challenging). An illustration is given in Figure~\ref{fig:deconv_demo}.

@@

R and Python code
for reproducing all figures and results in this paper can be found at
\url{https://github.com/cmu-delphi/stat-sci-nowcast}.

@@

The latent setting exhibits another degree of diversity within itself. In our
work, we target symptomatic COVID-19 infections, which, to be perfectly clear,
is a latent time series. Another example along similar lines is
\citet{Goldstein:2009}, who estimate influenza infection incidence via Bayesian
deconvolution of mortality data. Meanwhile, other authors might view inferring 
latent infections as just a stepping stone toward ultimately estimating the 
instantaneous reproductive number $R_t$, a key epidemic parameter. Important 
contributions to the methodology on real-time estimation of $R_t$ include: 
%\citet{Wallinga:2004}, who propose a forward-looking approach (in fact, to
%estimate a related but slightly different quantity known as the case
%reproduction number) based on an estimate of the serial interval;
\citet{Bettencourt:2008}, who use a local approximation to the SIR model, and 
\citet{Cori:2013, Thompson:2019}, who use a discretization of the renewal
equation within a Bayesian framework. For a thorough review and comparison of
these methods, see \citet{Gostic:2020}. The latter paper also discusses in some
detail the importance of properly modeling the delay between infection onset and
case report, and the issue of right truncation, which, as we will see, are
central issues in our paper as well.

The aforementioned methods have been applied and extended to build systems for
real-time $R_t$ nowcasting during the COVID-19 pandemic by \citet{Abbott:2020,
  rtlive, Chitwood:2021}. A key difference between these approaches and ours is 
that they infer infections through forward-filling: loosely speaking, they
convolve forward a candidate estimate of infections, obtain feedback by
comparing the result to measured cases, and iterate to refine estimates. This
can be effective given accurate prior knowledge, but of course it can be hard to
judge the accuracy of prior knowledge in practice. We take a more flexible,
non-model-based approach and estimate infections via direct deconvolution. We
also focus on fusing in auxiliary sources of information in order to improve
real-time accuracy and robustness. We remark that, if estimates of $R_t$ were
desired, then these could certainly be inferred as a by-product of our infection
nowcasts.

Finally, deconvolution has been extensively studied for many years in many
fields, notably signal and image processing, where deconvolution is sometimes
called deblurring. As an inverse problem, deconvolution is ill-posed in settings
in which the convolution operator is not known exactly or observations are made
with noise \citep{Oppenheim:2017}. Approaches to overcome this traditionally
involve regularization, as in the classical Wiener deconvolution
\citep{Wiener:1964}, which stabilizes the inversion using an estimated
signal-to-noise ratio. Alternative approaches employ familiar regularization
techniques such as $\ell_1$ and $\ell_2$ penalities \citep{Taylor:1979,
  Debeye:1990}. Most related to our paper is deconvolution using total variation 
regularization, first proposed by \citet{Rudin:1994}, and now a central tool in
signal and image processing.

@@

Roughly speaking, we study these particular indicators (ordered roughly from
``late'' to ``early'') because conceptually they reflect data measurements that
would be made at some period of time in between infection onset and case report
to a public health authority, and therefore would be relevant in inferring
latent infection rates. More information on these indicators and their
underlying data sources is given in \citet{Reinhart:2021}. For more information
on CTIS in particular, see \citet{Salomon:2021}; and for a study of how these
and similar indicators can improve COVID-19 forecasting, see
\citet{McDonald:2021}.

@@

In this paper, we consider (as described in Section~\ref{sec:sensor_fusion})
various methods for combining these estimates into a single estimate of the  
infection rate, which we will call \emph{sensor fusion} methods. Broadly
speaking, sensor fusion is a form of ensembling, which is ubiquitous in in
predictive modeling in statistics and machine learning, as it can often help
improve both accuracy and robustness. In our particular application, the sensors
themselves are constructed from data streams operating outside of traditional
public health reporting, which itself contributes an additional important angle
in terms of robustness. % in the estimation of latent infection rates (as these 
% are tied to  public health reports via a convolutional model).

@@

When we say above that nowcasts are made using data that would have been
available \emph{as of} a given nowcast date $t$, we mean that we adhere not to
only the real-time availability (latency) of signals at $t$, but also the
\emph{version} of the data published at $t$---simply put, imagine that we
``rewind'' the clock to time $t$ and query the API to receive the data that
would have been returned then. This is possible becausse the COVIDcast API
records and provides access to all historical versions of data, as described in
\citet{Reinhart:2021}. As epidemic data is often subject to revision, if we
train and evaluate models on ``finalized'' data (that would have been available
only at a much later time point) then this can lead to inaccurate conclusions
about real-time model performance; see, e.g., \citet{McDonald:2021}.

Further, it is worth noting that reported case data from JHU is available at a
1-day lag, and we assume that there is at least another 1-day lag between
symptom onset and case report (explained in
Section~\ref{sec:delay_distribution}). Hence through real-time deconvolution
alone we would be able to make nowcasts at a 2-day lag at the earliest. Making
nowcasts at a 1-day lag is possible with sensor fusion, using auxiliary signals 
with 1-day latency (explained in Section~\ref{sec:leverage_aux}). In this sense,
sensor fusion is able to improve not only accuracy, but also latency, and buys
us 1 extra day.

@@

We produce nowcasts at the county resolution, but for computational purposes, we
restrict our attention to the 200 U.S.\ counties with the highest
population. We additionally produce estimates for each of the 50 U.S.\ states. 
(Some of the methodology that we use for sensor fusion requires a geographical
hierarchy, thus using the remaining $\approx$ 3000 U.S.\ counties we aggregate
these within each state to create ``rest-of-state'' jurisdictions, and make
estimates for these as well, for the purposes or maintaining such a hierachy.)

@@

Lastly, under an assumption that the confounding acts as a multiplicative bias
that changes slowly over time, our real-time infection rate
estimates---themselves subject to confounding, as explained above---can be
post-processed to derive real-time \emph{approximately unconfounded} estimates
of $R_t$. This is also described in the discussion.

@@

Assumption~\ref{asm:geo_invar} is the most noteworthy and troublesome. We do
\emph{not} believe it to be true that different locations actually have
identical patterns of delay between infections and case reports; conversely, we
expect there to be a considerable amount of variability between locations in
this regard. While we do allow the delay distribution to change over time (see
Figure~\ref{fig:line_list_time} for evidence for the importance of this), we
consider Assumption~\ref{asm:geo_invar} to be a weakness of our work. However,
the \emph{data is simply not there} in the CDC line list to warrant
location-specific estimation of the delay distribution (see
Figure~\ref{fig:line_list_state}), thus we resort to estimating a nation-wide
delay distribution.

@@

\begin{figure}[tb]
\centering	
\includegraphics[width=0.95\linewidth]{./figures/right_truncation_illustration.pdf}
\caption{Illustration of right truncation with a delay distribution of length 3
  (which is taken to be stationary for simplicity). At the nowcast time $t$,
  some ``part'' of the latent signal $x_t$ will appear in $y_{t+1},y_{t+2}$;
  likewise, some ``part'' of $x_{t-1}$ will appear in $y_{t+1}$.}   
\label{fig:right_truncation}
\end{figure}

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=.315\linewidth]{./figures/ny_tf.pdf}
\includegraphics[width=.315\linewidth]{./figures/ny_ntf.pdf}
\includegraphics[width=.315\linewidth]{./figures/ny_tapered_ntf.pdf}
\caption{Comparison of boundary behavior for real-time deconvolution in New 
  York, displayed for a sample of different nowcast dates (where each colored
  curve traces out the deconvolution estimates for a different nowcast
  date). The black dashed line indicates finalized infections, estimated roughly
  three months after June 1, 2021.}  
\label{fig:boundary_comparison}
\end{figure*}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.825\linewidth]{./figures/tapered_smoothing_03_small_square.pdf}
\caption{Comparing regularization approaches by MAE for nowcasting (the shaded
  bands here and henceforth, in all MAE figures, correspond to 95\% bootstrap
  confidence intervals.) Both approaches for additional regularization give a
  huge improvement on trend filtering. The biggest improvement comes from
  combining the two approaches.}
	\label{fig:extra_reg}
\end{figure}

@@

The instantaneous reproductive number $R_t$, the average number of secondary
infections at time $t$ generated from a primary infection in the past, is a
useful and interpretable parameter that reflects the dynamics of epidemic growth
in a population. In the SIR model, the instantaneous reproduction number $R_t$
and growth rate $r_t$ at time $t$ obey the following relationship:
\[
R_t \approx 1 + \frac{r_t}{\gamma}, 
\] 
where $\gamma$ denotes the recovery rate in the SIR model. While this is
well-known in the literature on mathematical modeling of epidemics (and is exact
under local exponential growth; see, e.g., \citet{Wallinga:2007}), its use in
the presence of confounding seems to be underexplored and potentially
undervalued. If $I_t$ denotes the number of new infections at $t$, then using a
simple discrete difference approximation to $r_t$ leads to: 
\[
R_t \approx 1 + \frac{1}{\gamma}\bigg( \frac{I_{t+1}}{I_t} - 1 \bigg). 
\]
A similar though not identical approximation is given in
\citet{Bettencourt:2008}, where $I_{t+1}/I_t-1$ is replaced by
$\log(I_{t+1}/I_t)$. Critically, incident infections only enter right-hand side
above as a \emph{ratio} of values adjacent in time, and thus if we are only able
to estimate this up to an unknown multiplicative factor (due to confounding),
then this factor approximately cancels in the ratio as long as it is slowly
varying in time. In slightly more detail (and for simplicity, considering just a
single location), suppose as before that a fraction $a_t$ of infections go
untested at time $t$. Then $I_t = b_t x_t$ where $x_t$ is the number of new
infections at time $t$ that show up in case reports (i.e., the focus of this
paper) and $b_t = 1/(1-a_t)$. From the previous display,   
\[
R_t \approx 1 + \frac{1}{\gamma}\bigg( \frac{b_{t+1} x_{t+1}}{b_t x_t} - 1
\bigg) \approx 1 + \frac{1}{\gamma}\bigg( \frac{x_{t+1}}{x_t} - 1 \bigg),
\]
where the last approximation is motivated by an additional assumption the 
untested fraction varies slowly over time (so $b_{t+1}/b_t \approx
1$). This shows that estimates of $x_t$ can produce \emph{approximately
  unconfounded} estimates of $R_t$, even though $x_t$ is itself confounded
due to a lack of universal testing. This is true both in the retrospective and 
real-time sense, and will be the topic of future study.

@@

% \section{Log-Change-Ratio Approximation for $R_t$} 
% \label{app:log_change_ratio}

% We consider the standard susceptible-infected-recovered (SIR) model
% \citep{Kermack:1927} with transmission rate $\beta$ and recovery rate
% $\gamma$. Let $S(t),I(t),R(t)$ denote the fraction of susceptible, infected,
% recovered individuals, respectively, at time $t$, in some underlying
% population. Then the SIR model evolves according to the continuous-time dynamics
% (subject to some initial conditions):
% \begin{align*}
% S'(t) &= -\beta S(t) I(t) \\
% I'(t) &= \beta S(t) I(t) - \gamma I(t) \\
% R'(t) &= \gamma I(t).
% \end{align*}

% The instantaneous reproduction number in this model is defined as $R_t = 
% S(t) \beta / \gamma$. We can interpret the $R_t$ as being determined by 
% the slope of a natural first-order approximation to the curve $\log I(t)$ at
% time $t$. That is, looking back at the differential equation for $I(t)$, if we
% forget about the fact that $S(t)$ actually depends on $t$, this ``looks like'' 
% $I'(t)$ is a ``constant'' times $I(t)$ (that ``constant'' being $\beta S(t) -
% \gamma$), which suggests that for $t$ near a point $t_0$,   
% \begin{align}
% \nonumber
% I(t) &\approx I(t_0) \exp\big( (\beta S(t) - \gamma) (t-t_0) \big) \\ 
% \label{eq:inf_approx}
% &= I(t_0) \exp\big(\gamma (R_t - 1) (t-t_0)\big).
% \end{align}

% Now let $T(t)$ denote the total fraction of individuals who have ever been
% infected at time $t$ (note, $T(t) = 1-S(t)$). Then this evolves according to the 
% dynamics: 
% $$
% T'(t) = \beta S(t) I(t).
% $$
% Consider the change in total infecteds from time $t$ to $t+1$, also called the
% \emph{incidence proportion} over this period, denoted $\Delta T(t+1) = T(t+1) - 
% T(t)$. Using this as a discrete approximation to the derivative (backward
% difference) of the total infecteds curve, 
% \begin{align*}
% \Delta T(t+1) &\approx \beta S(t+1) I(t+1) \\
% &\approx \beta S(t) I(t) \exp\big(\gamma (R_t - 1)\big),
% \end{align*}
% where in the second line we used the local-linear approximation to log-infecteds
% curve from \eqref{eq:inf_approx}, as well as a local-constant approximation to
% the susceptible curve (which underlies this local-linear approximation in the
% first place). By the same logic,  
% $$
% \Delta T(t) \approx \beta S(t) I(t),
% $$
% and thus 
% $$
% \frac{\Delta T(t+1)}{\Delta T(t)} \approx \exp\big(\gamma (R_t - 1)\big). 
% $$
% In other words, from the ratio $\Delta T(t+1) / \Delta T(t)$, we can estimate
% $R_t$, provided that we already know (or can estimate) $\gamma$. This
% establishes the claim in \eqref{eq:log_change_ratio} by taking logs of each
% side, and noting that if $N_t$ is the number of new infections at each $t$, then
% $N_{t+1} / N_t = \Delta T(t+1) / \Delta T(t)$ (because the total number of
% individuals in the population cancels in the ratio).
\end{appendix}
